---
title: "Interactive Confusion Simplexes"
subtitle: "Isometric 3D projections of confusion matrices"
knit: (function(input_file, encoding) {
  out_dir <- 'docs';
  rmarkdown::render(input_file,
    encoding=encoding,
    output_file=file.path(dirname(input_file), out_dir, 'ConfusionSimplex.html'))})
author: "David Lovell"
date: "January 2023"
output: html_document
---

```{r setup, include=FALSE}
# Start with a clean environment
rm(list=ls())

knitr::opts_chunk$set(echo = TRUE, comment = NA)
library(tidyverse)
library(magrittr)
library(extraDistr)        # For beta-binomial distribution functions
library(plotly)            # for 3-D visualisation
library(RColorBrewer)
library(patchwork)         # for arranging figures
library(scales)
library(ggridges)

c.low  <- rgb(178/255,24/255,43/255)
```

```{r function-makeAllWeakCompositions, echo=FALSE}
makeAllWeakCompositions <- function(n,k){
  # Initialise the matrix that will hold all compositions
  composition <- matrix(data=0, nrow=choose(n+k-1,k-1), ncol=k)
  
  composition[1,k]  <- n # Set the first composition (0,...,0,n)
  current.row       <- 1 # Set the current row to the first row
  last.nonzero      <- k # The last non-zero element of the current row is in position k
  
  # While the first element of the current row is less than n... 
  while(composition[current.row,1] < n){
    # generate the next row
    next.row <- current.row + 1
    # copy the current row into the next row
    composition[next.row,] <- composition[current.row,]
    # turn    a b ...   y   z 0 0 ...   0
    #                       ^ last
    # into    a b ... (y+1) 0 0 0 ... (z-1)
    
    last.nonzero                            <- max(which(composition[next.row,] > 0))
    z                                       <- composition[current.row, last.nonzero]
    composition[next.row, last.nonzero - 1] <- composition[current.row, last.nonzero - 1] + 1
    composition[next.row, last.nonzero    ] <- 0
    composition[next.row, k               ] <- z - 1
    current.row                             <- next.row
  }
  return(composition)
}
```


```{r function-MCC, echo=FALSE}
MCC <- function(TP, FP, FN, TN){
  MCC <- (TP*TN - FP*FN)/sqrt((TP+FP)*(TP+FN)*(TN+FP)*(TN+FN))
  MCC <- ifelse(
    is.finite(MCC), MCC,
    ifelse(
      (TP == 0 & FP == 0) | (TP == 0 & FN == 0) | (TN == 0 & FP == 0) | (TN == 0 & FN == 0),
      0,
      ifelse(
        (TP > 0 & FP + FN + TN == 0) | (TN > 0 & FP + FN + TP == 0),
        +1,
        -1)
    )
  )
  return(MCC)
}
```

```{r function-zdiv, echo=FALSE}
zdiv <- function(x,y){
  ifelse(
    x==0, 
    ifelse(y==0,NA,0),
    x/(x+y)
    )
} 
```

```{r function-make.confmat, echo=FALSE}
make.confmat <- function(n) {
  # This matrix is used to project the four dimensional confusion matrix into three dimensions
  project3d <- matrix(
    c(
        0 ,   0 ,   1, # TP
        0 ,   1 ,   0, # FP
        1 ,   0 ,   0, # FN
      -1/3, -1/3, -1/3 # TN
    ), byrow = TRUE, nrow=4
  )
  
  makeAllWeakCompositions(n,4) -> abcd              # All confusion matrices of size n
  colnames(abcd) <- c("TP", "FP", "FN", "TN")       # with columns named after confusion matrix elements
  abcd %*% project3d           -> xyz               # ...projected into 3D
  colnames(xyz)  <- c("x", "y", "z")                # with columns named after the three dimensions
  bind_cols(as_tibble(abcd), as_tibble(xyz)) %>%    # ...bound side by side
    mutate(                                         # and augmented with...
      text       =sprintf("%2d %2d\n%2d %2d", TP,FP,FN,TN), # Label for plotly
      Pos        =TP+FN,                            # Number of actual positives
      Neg        =FP+TN,                            # Number of actual negatives
      TPR        =TP/Pos,                           # True Positive Rate
      FPR        =FP/Neg,                           # False Positive Rate
      PLR        =TPR/FPR,                          # Positive Likelihood Ratio (LR+)
      TNR        =TN/Neg,                           # True Negative Rate
      FNR        =FN/Pos,                           # False Negative Rate
      NLR        =FNR/TNR,                          # Negative Likelihood Ratio (LR-)
      DOR        =PLR/NLR,                          # Diagnostic Odds Ratio
      prior.O    =Pos/Neg,                          # Prior odds of actual class being X
      prior.P    =zdiv(Pos,Neg),                    # Prior prob of actual class being X
      post.O     =TP/FP,                            # Posterior odds that actual class is X
      post.P     =zdiv(TP,FP),                      # Posterior probability that actual class is X 
      prior.O.n  =Neg/Pos,                          # Prior odds of actual class NOT being X
      prior.P.n  =zdiv(Neg,Pos),                    # Prior prob of actual class NOT being X
      post.O.n   =TN/FN,                            # Posterior odds that actual class is NOT X 
      post.P.n   =zdiv(TN,FN),                      # Posterior probability that actual class is NOT X
      MCC        =MCC(TP,FP,FN,TN),                 # Matthews correlation coefficient
      logDOR     =log(DOR),                         # log of DOR
      #slogDOR    =logDOR/log((Pos-1)*(Neg-1)),      # scaled log of DOR
      J          =TPR + TNR - 1,                    # Youden's J, Balanced Accuracy
      Acc        =(TP+TN)/(Pos + Neg),              # Accuracy
      F1         =2*TP / (2*TP + FP + FN),          # F1
      Markedness =post.P + post.P.n - 1,            # Markedness
      g.mean     =sqrt(TPR * TNR),                  # Geometric mean
      Prev.Thresh=sqrt(FPR)/(sqrt(TPR)+sqrt(FPR)),  # Prevalence threshold
      Threat.Scr =TP / (TP + FN + FP),              # Threat score
      Fowlkes.M  =sqrt(post.P * TPR),               # Fowlkes-Mallows index
    ) 
  }
```

```{r make.confmats, echo=FALSE, warning=FALSE}
# Make some sets of confusion matrices that we will use later
make.confmat(100) -> confmat.100
```

```{r function-plot.simplex, echo=FALSE}
# This convenience function plots the 3D projection of the confusion matrices in confmat
# coloured by metric
# for three different values of Pos
plot.simplex <- function(confmat, metric="MCC", Pos.vals=c(20, 50, 90)){
  # Pick out the extremes of the confusion simplex
  N <- max(confmat$TP)
  extremes <- filter(confmat, pmax(TP,FP,FN,TN) == N)
  edges    <- rbind(
    extremes,
    extremes[c(1,3,2,4),]
  )
  
  # Set up a colour scale that works for finite values of the chosen metric
  select(confmat, all_of(metric)) %>%
    filter(is.finite(.[[metric]])) %>% range() -> metric.range

  marker <- list(
    color=as.formula(paste("~", metric)),
    cmin=metric.range[1],
    cmax=metric.range[2],
    colorscale=list(
      seq(from=0, to=1, len=11), brewer.pal(11, "RdBu")
    ),
    colorbar=list(
      x=0, y=0.5, outlinewidth=0, len=0.5, title=list(text=metric)
    ),
    size=2, showscale=TRUE
  )
  
  # Set up the axes and annotations
  scene <- list(
    xaxis = list(visible=FALSE),
    yaxis = list(visible=FALSE),
    zaxis = list(visible=FALSE),
    annotations=list(
      list(x=N, y=000, z=000, showarrow=FALSE, text="FN", yanchor="top"),
      list(x=000, y=N, z=000, showarrow=FALSE, text="FP", yanchor="top"),
      list(x=000, y=000, z=N, showarrow=FALSE, text="TP", yanchor="bottom"),
      list(x=-N/3, y=-N/3, z=-N/3, showarrow=FALSE, text="TN", yanchor="top")
    )
  )
  
  # Create the plot
  plot_ly(
    # plot the finite values of the slice where Pos == Pos.vals[1]
    data=
      select(confmat, x, y, z, Pos, text, all_of(metric)) %>% 
      filter(Pos == Pos.vals[1], is.finite(.[[metric]])),
    type="scatter3d", mode="markers", name=sprintf("Pos==%d", Pos.vals[1]),
    x = ~x, y = ~y, z = ~z, marker=marker, text=~text, hoverinfo="text"
  ) %>% 
    add_trace(
    # plot the finite values of the slice where Pos == Pos.vals[2]
      data=select(confmat, x, y, z, Pos, text, all_of(metric)) %>% 
        filter(Pos == Pos.vals[2], is.finite(.[[metric]])),
      type="scatter3d", mode="markers", name=sprintf("Pos==%d", Pos.vals[2]),
      x = ~x, y = ~y, z = ~z, marker=marker, text=~text, hoverinfo="text"
    ) %>% 
    add_trace(
    # plot the finite values of the slice where Pos == Pos.vals[3]
      data=select(confmat, x, y, z, Pos, text, all_of(metric)) %>% 
        filter(Pos == Pos.vals[3], is.finite(.[[metric]])),
      type="scatter3d", mode="markers", name=sprintf("Pos==%d", Pos.vals[3]),
      x = ~x, y = ~y, z = ~z, marker=marker, text=~text, hoverinfo="text"
    ) %>% 
    # plot the edges of the confusion simplex
    add_trace(
      data=edges,  name="edges", mode="markers+lines",
      x=~x, y=~y, z=~z,  marker=list(showscale=FALSE), line=list(color="#DDDDDD")
    ) %>%
    # allow points to be interactively identified
    layout(
      hoverlabel=list(font=list(family = "monospace")),
      scene=scene
    )
}
```

## Confusion simplex coloured by Accuracy


```{r echo=FALSE}
plot.simplex(confmat.100, metric="Acc")
```

* Mouse over the tetrahedron, then click and drag to change its orientation.
* Click on the text `Pos==20` to toggle that slice of the confusion matrix.

## Confusion simplex coloured by Balanced Accuracy


```{r echo=FALSE}
plot.simplex(confmat.100, metric="J")
```

* Mouse over the tetrahedron, then click and drag to change its orientation.
* Click on the text `Pos==20` to toggle that slice of the confusion matrix.

## Confusion simplex coloured by F1 Score


```{r echo=FALSE}
plot.simplex(confmat.100, metric="F1")
```

* Mouse over the tetrahedron, then click and drag to change its orientation.
* Click on the text `Pos==20` to toggle that slice of the confusion matrix.

## Confusion simplex coloured by Matthews Correlation Coefficient

```{r echo=FALSE}
plot.simplex(confmat.100, metric="MCC")
```

* Mouse over the tetrahedron, then click and drag to change its orientation.
* Click on the text `Pos==20` to toggle that slice of the confusion matrix.


## About

These [interactive plotly visualisations](https://plotly.com/) show 3D projections of binary confusion matrices of size $N=100$.

* Each plot depicts the value of a different confusion matrix performance metric.
* Each point corresponds to a unique confusion matrix and is coloured by the value of that metric. 
* For reference, we label the four extreme points corresponding to all True Positives, ($\mathrm{TP}=100$), all False Negatives ($\mathrm{FN}=100$), etc., and connect those vertices to give an impression of the regular tetrahedral lattice (i.e., the 3-simplex) of the projected points.
* In total, there are $\binom{100+4-1}{4-1}=176\,851$ different binary confusion matrices with of size 100.
* Rather than show all these, we have taken three slices through the lattice: from back to front, the rectangular lattices of points correspond to confusion matrices where $\mathrm{Pos} = 20, 50, 90$, respectively.
* If you rotate the plots so that $\mathrm{TP}$ is at the top, $\mathrm{FN}$ is at the back left and $\mathrm{FP}$ at the back right, the axes of these slices will be oriented like those of conventional ROC curves:
  * $(\mathrm{TN},\mathrm{TP})$ are maximum at the top-left and minimum at the bottom-right;
  * the diagonals from bottom-left to top-right correspond to confusion matrices from random classification.



